Distinguishing multiple-merger from Kingman coalescence using two-site frequency spectra

### 2018-10-06 manuscript/manuscript.tex
Maryn suggestions (part 1).

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are simplified mathematical models, which connect unobserved quantities such as the population size to observable features of genetic data.
However, populations are complicated and, moreover, vary in their complications.
A simple model cannot capture the processes governing every species' evolution, and fitting a misspecified model will generate misleading inferences.
It is therefore crucial to understand the limits of our models and to be able to assess when a model is appropriate for a particular data set.

One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982, Kingman1982a, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that allow for both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model can be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
Under a variety of conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygosity along a chromosome (e.g., \cite{LiDurbin2011}).

A serious problem for this class of inference methods is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
For example, one of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For instance, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly} rather than inversely proportional to the population size.
It follows that the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.

@@

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} \autocite{Pitman1999, Sagitov1999, DonnellyKurtz1999,} (reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The even more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
These models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
and purifying selection at sufficiently many sites \autocite{SegerEtAl2010, NicolaisenDesai2012, GoodEtAl2014}.

In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.

It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{VittiEtAl2013}).
These methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
Genomic regions that are outliers from this distribution are presumed to be under selection.
In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
\cite{BirknerEtAl2013, BlathEtAl2016, SpenceEtAl2016} derived methods for computing the expected site frequency spectrum of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the site frequency spectrum is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.

However, existing methods have limitations for distinguishing multiple mergers from general models of population-size change.
The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but detecting it in data requires knowledge of the ancestral allele at each site.
High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS at high frequencies.

@@

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without any information about their relative locations in the genome.

In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
We define the 2-SFS,
$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
This object has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and \cite{Xie2011a} in a model with selection.
We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies. (For non-recombining sites, the 2-SFS is independent of the distance and so we will suppress the $d$ in our notation.)

In the limit of low per-site mutation rate ($\mu\to0$), and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the genealogical branch length distribution by
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
    \label{eq:expected_2sfs}
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, varies from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

In the following, we demonstrate that the positive associations between mutations at different frequencies distinguish the multiple-merger from the Kingman coalescent and that this distinction
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.
To do so, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
Finally, we demonstrate the use of $\fpmi$ in a coalescent model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We are interested in quantifying the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
To do so, we define the \textit{frequency pointwise mutual information} as
\begin{align}
    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
    \label{eq:fpmi}
\end{align}
\Fig{fig:schematic} shows the steps to compute $\fpmi$ from a sample of sequences.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/schematic.pdf}
\caption{Summarizing population genetic data with frequency pointwise information. First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations. Then, for a given distance, $d$, the 2-SFS gives the number of pairs of sites with each combination of minor allele frequencies. Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
 \label{fig:schematic}}
\end{figure}

In information theory, the pointwise mutual information (PMI) of a pair of random variables, $X$ and $Y$, is a transformation of the probability mass function, $p(x,y)$, given by
\begin{align}
    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)}
    \label{eq:pmi}
\end{align}
where $p(x) = \sum_y p(x,y)$ and vice versa \autocite{ChurchHanks1990}.
PMI measures the change in the probability that $X=x$ given knowledge that $Y=y$.
When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
On the other hand, when $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and conversely for $\pmi(x,y) < 0$.
(The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.)

In our setting, the 2-SFS may be interpreted as a joint probability mass function over minor allele counts.
Given a coalescent model, the minor allele count at an arbitrary site is a random variable over $\{0,\ldots,\floor{n/2}\}$ with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
Similarly, the minor allele counts at a pair of sites separated by $d$ bases is a pair of random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of minor allele counts.

@@

We claim that the second benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous version.
On the other hand, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.

>>>>

\section*{Introduction}
The genetic diversity of a population reflects its demographic and evolutionary history.
Learning about this history from contemporary genetic data is the domain of modern population genetics (see~\cite{Hahn2018a}).
The fundamental tools of the trade are simplified mathematical models, which connect unobserved quantities such as the population size to observable features of genetic data.
However, populations are complicated and, moreover, vary in their complications.
A simple model cannot capture the processes governing every species' evolution, and will generate misleading inferences if misspecified.
It is therefore crucial to understand the limits of our models and to assess when a model is appropriate for a particular data set.

One of the most widely used models is the Kingman coalescent~\autocite{Kingman1982, Kingman1982a, Hudson1983, Tajima1983}.
The Kingman coalescent is a stochastic process that generates gene genealogies: trees representing the patterns of shared ancestry of sampled individuals.
Inference methods use these genealogies as latent variables linking demographic parameters to genetic data~\autocite{RosenbergNordborg2002}.
The Kingman coalescent has a number of convenient properties that facilitate both analytical calculations (e.g.,~\cite{Tajima1989}) and efficient stochastic simulations (e.g.,~\cite{Hudson2002}): tree topologies are independent of waiting times; waiting times are generated by a Markov process; and neutral mutations are modeled as a Poisson process conditionally independent of the tree.
Moreover, the model can be extended to study a variety of biological phenomena including recombination, population structure, and variation in sex ratios or ploidy (see generally~\cite{Wakeley2009}).

An important application of the Kingman coalescent is inferring historical population sizes from genetic data~\autocite{SchraiberAkey2015}.
In its simplest form, the model has a single parameter, the coalescent rate, which determines the branch lengths of genealogies~\autocite{Kingman1982}.
Under many conditions, the coalescent rate is inversely proportional to the population size~\autocite{Kingman1982a}.
Accordingly, a growing or shrinking population may be modeled by a time-varying rate~\autocite{GriffithsTavare1994, GriffithsTavare1998}.
Patterns of genetic diversity depend on the ratio of the coalescent rate to other evolutionary rate parameters.
For example, the \emph{site frequency spectrum} (SFS)---the number of mutations segregating at different frequencies in a sample---is determined by the ratio of the mutation rate to the (time-varying) coalescent rate.
Kingman coalescent--based inference methods solve the inverse problem of determining the population size history that best explains particular features of the data, such as the SFS (e.g., \cite{BhaskarEtAl2015}) or variations in heterozygosity along a chromosome (e.g., \cite{LiDurbin2011}).

A serious problem for this class of inference methods is that different models of evolution generate different relationships between historical population sizes and genetic diversity.
For example, one of the basic assumptions of the Kingman coalescent is that natural selection is negligible in determining the distribution of genealogies.
When this assumption is violated, Kingman-based inference methods are misspecified.
For instance, when a beneficial mutation increases rapidly in frequency, it distorts the genealogies at nearby sites (see e.g., \cite{CoopRalph2012}).
If these ``selective sweeps'' occur regularly, they may be the dominant factor determining the distribution of genealogies.
In this case, the average coalescent rate is proportional to the number of beneficial mutations introduced per generation, which is itself \emph{directly}, rather than inversely, proportional to the population size.
It follows that the relationship between the population size and the expected number of neutral mutations in a sample is inverted: larger populations will be less diverse than smaller populations.

@@

An important extension of the Kingman coalescent is a family of models known as \textit{multiple merger coalescents} (\cite{Pitman1999, Sagitov1999, DonnellyKurtz1999,}; reviewed in \cite{Eldon2016}), which arise in a variety of contexts both with and without selection.
Whereas in the Kingman coalescent lineages may coalesce only pairwise, multiple merger coalescents permit more than two lineages to coalesce in a single event.
The more general class of simultaneous multiple merger coalescents \autocite{Schweinsberg2000, MohleSagitov2001, Sagitov2003} permits more than one distinct multiple merger event at the same time.
Multiple merger and simultaneous multiple merger models are relevant for species with
``sweepstakes'' reproductive events \autocite{EldonWakeley2006, SargsyanWakeley2008},
fat-tailed offspring number distributions \autocite{Schweinsberg2003},
recurring selective sweeps at linked sites \autocite{DurrettSchweinsberg2005, CoopRalph2012},
rapid adaptation \autocite{NeherHallatschek2013, DesaiEtAl2013},
and purifying selection at sufficiently many sites \autocite{SegerEtAl2010, NicolaisenDesai2012, GoodEtAl2014}.

In each of these contexts, the coalescent timescale is not necessarily proportional to the population size.
For example, with fat-tailed offspring distributions the rate of coalescence is a power law in the population size \autocite{Schweinsberg2003}, while with linked sweeps it is determined by the rate of linked sweeps, as described above \autocite{DurrettSchweinsberg2005}.
In these settings, interpreting the level of genetic diversity in terms of an ``effective population size'' is misleading and inferences based on the Kingman coalescent may be qualitatively incorrect.

It is therefore important to determine whether the Kingman model is appropriate for a given data set before performing demographic inference.
This task is distinct from ``selection scan'' methods designed to detect particular regions of the genome that are under selection (see~\cite{VittiEtAl2013}).
Selection scan methods typically assume that most of the genome is evolving neutrally and that the genome-wide distribution of summary statistics reflects demographic factors.
Genomic regions that are outliers from this distribution are presumed to be under selection.
In contrast, we are interested in detecting when the genome-wide background \emph{is not} well-modeled by the Kingman coalescent.

One approach to identifying multiple mergers in genomic data is to use the SFS as a summary statistic.
To this end, \cite{BirknerEtAl2013, BlathEtAl2016} and \cite{SpenceEtAl2016} derived methods for computing the expected SFS of (simultaneous-)multiple merger coalescents.
\cite{EldonEtAl2015} showed that it is possible to distinguish between a multiple merger coalescent of the beta family and the Kingman coalescent with exponential growth using the SFS.
\cite{RodelspergerEtAl2014} detected widespread linked selection in the nematode \textit{Pristionchus pacificus} by demonstrating that the SFS is non-monotonic, a signature of multiple mergers \autocite{NeherHallatschek2013, BirknerEtAl2013}.

However, existing methods are limited in their ability to distinguish multiple mergers from general models of population-size change.
The primary signature of multiple mergers in the SFS is an overabundance of low-frequency mutations relative to the Kingman expectation, which is also the signature of population growth.
\cite{EldonEtAl2015} were able to reject exponential growth in favor of multiple mergers with sufficient data, but a more flexible model of growth may be able to fit the multiple mergers SFS (see \cite{MyersEtAl2008, BhaskarSong2014}).
The non-mononotic SFS identified by \cite{RodelspergerEtAl2014} is a more robust signature of multiple mergers, but identifying that the SFS increases at high frequencies requires knowledge of the ancestral allele at each site.
High-frequency mutations are typically much rarer than low-frequency mutations, so misidentifying even a small fraction of ancestral alleles can generate a non-monotonic SFS.

@@

Following the notation of \cite{Fu1995}, the site frequency spectrum of a sample of $n$ haploid genomes is $\left\{ \xi_i : 1 \leq i < n \right\}$, where $\xi_i$ is the fraction of sites containing a mutation with derived allele count $i$ in the sample.
In many cases, the ancestral allele is unknown and so the allele in $i$ samples and the complementary allele in $n-i$ samples are indistinguishable.
Therefore, we will mostly consider the \textit{folded} site frequency spectrum $\left\{ \eta_i = \xi_i + (1-\delta_{i,n-i}) \xi_{n-i}: 1 \leq i \leq \floor{n/2} \right\}$, where $\delta_{k,k'}$ is the Kronecker delta. %, where $\eta_i = \xi_i + \delta_{n-2i} \xi_{n-i}$.
The SFS and folded SFS can be calculated from a set of single nucleotide polymorphisms (SNPs) without knowledge of the physical locations of the SNPs.

In contrast, the 2-SFS is a statistic of \textit{pairs} of sites.
We define the 2-SFS,
$\left\{ \xi_{ij}(d) : d > 0; 1 \leq i, j < n\right\}$,
as the fraction of pairs of sites separated by $d$ bases for which there is a mutation with derived allele count $i$ at one site and a second mutation with derived allele count $j$ at the other site.
(Note that $\xi_{ij}(d) = \xi_{ji}(d)$ by symmetry.)
The 2-SFS has been studied for non-recombining sites by \cite{FerrettiEtAl2018} in a neutral model and by \cite{Xie2011a} in a model with selection.
We define the folded 2-SFS, $\eta_{ij}(d)$, by analogy to the folded SFS,  categorizing pairs of sites by their minor allele frequencies. (For non-recombining sites, the 2-SFS is independent of the distance and so we will suppress the $d$ in our notation.)

In the limit of low per-site mutation rate ($\mu\to0$) and no recombination, all polymorphic sites are bi-allelic and the expected SFS and 2-SFS are related to moments of the genealogical branch length distribution by
\begin{align}
    \E{\xi_i} &= \mu \E{\tau_i} \label{eq:expected_sfs} \\
    \E{\xi_{ij}} &= \mu^2 \E{\tau_i \tau_j},
    \label{eq:expected_2sfs}
\end{align}
where $\tau_i$ is the total length of branches subtending $i$ leaves of a gene genealogy and $\E{\cdot}$ represents the expectation over the distribution of gene genealogies defined by a coalescent model.
Thus, the SFS and 2-SFS depend on the distribution of coalescent times as well as the distribution of tree topologies.

@@

\cite{BirknerEtAl2013} extended Fu's calculation to a family of multiple merger coalescents called beta coalescents.
This one-parameter family interpolates between the Kingman coalescent and the Bolthausen-Sznitman coalescent as the parameter, $\alpha$, ranges from 2 to 1.
Beta coalescents arise in models with fat-tailed offspring distributions \autocite{Schweinsberg2003, SteinruckenEtAl2013}, and the Bolthausen-Sznitman coalescent is the limiting distribution of genealogies in populations that are rapidly adapting or experiencing extensive purifying selection \autocite{NeherHallatschek2013}.
The calculations of \cite{BirknerEtAl2013} show positive correlations between $\xi_i$ and $\xi_j$ for $j \not\in \{i, n-i\}$ (Figures 5 and 6).
Thus, unlike the standard Kingman coalescent, the beta coalescent can generate positive associations between mutations with different minor allele counts.

In the following, we demonstrate that the positive associations between mutations at different frequencies distinguish the multiple-merger from the Kingman coalescent and that this distinction
\begin{enumerate*}[label=(\roman*), before=\unskip{: }, itemjoin={{; }}, itemjoin*={{, and }}]
    \item applies to Kingman coalescents with time-varying coalescent rates
    \item is robust to recombination between the sites
    \item is also a feature of forward-time models with selection
    \item can form the basis of a model-checking procedure for demographic inference methods
\end{enumerate*}.
To demonstrate these four properties, we introduce a transformation of the 2-SFS that we call frequency pointwise mutual information ($\fpmi$).
We use a combination of numerical calculations and stochastic simulations to explore the properties of this statistic under different coalescent models.
Finally, we demonstrate the use of $\fpmi$ in a coalescent model-checking procedure on genomic diversity data from \textit{Drosophila melanogaster} \autocite{LackEtAl2015}.

@@

We aim to quantify the dependence between minor allele counts at a pair of sites, particularly for $i\neq j$.
To do so, we define the \textit{frequency pointwise mutual information} as
\begin{align}
    \fpmi(i,j;d) = \log \frac{\E{\eta_{i,j}(d)}}{\E{\eta_i} \E{\eta_j}}.
    \label{eq:fpmi}
\end{align}
\Fig{fig:schematic} shows the steps to compute $\fpmi$ from a sample of sequences.

\begin{figure}
\centering
\includegraphics[scale=1]{figures/schematic.pdf}
\caption{Summarizing population genetic data with frequency pointwise information ($\fpmi$). First, a sequence alignment is reduced to the minor allele frequencies and positions of segregating mutations. Then, for a given distance, $d$, the 2-SFS gives the number of pairs of sites with each combination of minor allele frequencies. Finally, $\fpmi$ is calculated from the 2-SFS according to \eq{eq:fpmi}.
 \label{fig:schematic}}
\end{figure}

In information theory, the pointwise mutual information (PMI) of a pair of random variables, $X$ and $Y$, is a transformation of the probability mass function, $p(x,y)$, given by
\begin{align}
    \pmi(x,y) = \log \frac{p(x,y)}{p(x)p(y)} = \log \frac{p(x|y)}{p(x)}
    \label{eq:pmi}
\end{align}
where $p(x) = \sum_y p(x,y)$ and similarly for $p(y)$ \autocite{ChurchHanks1990}.
PMI measures the change in the probability that $X=x$ given knowledge that $Y=y$.
When $X$ and $Y$ are independent, $\pmi(x,y) = 0$ for all $x,y$.
When $X$ and $Y$ are not independent, $\pmi(x,y) > 0$ implies that $p(x|y) > p(x)$, and $\pmi(x,y) < 0$ implies the converse.
(The expectation of PMI over the joint distribution of $X$ and $Y$ is known as the mutual information of $X$ and $Y$ \autocite{CoverThomas1991}.)

We can interpret the 2-SFS as a joint probability mass function over minor allele counts.
Given a coalescent model, the minor allele count at an arbitrary site is a random variable over $\{0,\ldots,\floor{n/2}\}$ with probability mass function $p(i) = \E{\eta_i}$, where we define $\eta_0$ to be the fraction of monomorphic sites.
Similarly, the minor allele counts at two sites separated by $d$ bases are random variables with joint probability mass function $p_d(i,j) = \E{\eta_{ij}(d)}$.
Comparing, \eq{eq:fpmi} with \eq{eq:pmi} shows that $\fpmi$ is the standard pointwise mutual information for pairs of minor allele counts.

@@

The primary benefit of computing $\fpmi$ from the 2-SFS is that it normalizes for distortions in the coalescent time distribution.
Population-size variation distorts the SFS by changing the distribution of waiting times.
However, the time-inhomogeneous Kingman coalescent generates the same distribution of tree topologies as the time-homogeneous coalescent.
In contrast, multiple mergers alter both the coalescent time distribution and the distribution of topologies.
These two effects are convolved in the SFS, making it difficult to distinguish between population growth and multiple mergers.
Our hope is that the $\fpmi$ reflects additional information about the distribution of genealogies, once distortions in coalescent times are accounted for.
That is, we expect $\fpmi$ to be sensitive to multiple mergers, but relatively unaffected by changes in the coalescent rate.
